Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-08-24 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
further minor additions and clarifications

Second, numerical errors can also be reduced by improving the accuracy in calculating the horizontal pressure gradient itself.  TF layers are usually implemented using a coordinate transformation onto a rectangular computational domain, which introduces metric terms into the equations of motion.  The techniques proposed by \citet{klemp2011} and \cite{zaengl2012} both involve the calculation of the horizontal pressure gradient in the physical domain \TODO{interpolate to z-levels in order to compute the horizontal pressure gradient}.  This gave them the flexibility to design more accurate horizontal pressure gradient discretizations using more appropriate stencils.

Despite their associated numerical errors, TF layers are in widespread operational use \citep{steppeler2003}.  They are attractive because their rectangular structure is simple to process by computer and link with parameterisations, boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}, and cell sizes remain almost constant \citep{jebens2011}. \TODO{clarify; constant cell sizes but variable spacing of vertical layers}

@@

Most implementations of terrain following layers use a coordinate system that makes the domain rectangular, but introduces metric terms into the equations of motion.  Instead, the model employed in this study uses Cartesian coordinates and unstructured grids \TODO{they're very structured!}.  By doing so, results from the same model can be compared between terrain following and cut cell grids without modifying the equation set or discretisation.

@@

\TODO{double check signs of all streamfunctions (in horizontal and TF advection test cases)}
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi\), along the edges from \(\Psi\) stored at cell vertices.  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{z - H}{\left( H - h \right)^2}, \nonumber \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

The test setup follows the specification by \cite{klemp2011}.  \TODO{how large is the domain? do I say this elsewhere? if not, say it here} The grid resolution is \(\Delta x = \Delta z^\star = \SI{500}{\meter}\).  All boundary conditions are no normal flow.

@@

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high.  \TODO{state grid spacing} The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.

>>>>

Second, numerical errors can also be reduced by improving the accuracy in calculating the horizontal pressure gradient itself.  TF layers are usually implemented using a coordinate transformation onto a rectangular computational domain, which introduces metric terms into the equations of motion.  The techniques proposed by \citet{klemp2011} and \cite{zaengl2012} both involve interpolation onto $z$-levels in order to calculate the horizontal pressure gradient.  This gave them the flexibility to design more accurate horizontal pressure gradient discretizations using more appropriate stencils.

Despite their associated numerical errors, TF layers are in widespread operational use \citep{steppeler2003}.  They are attractive because their rectangular structure is simple to process by computer and link with parameterisations, and boundary layer resolution can be increased with variable spacing of vertical layers \citep{schaer2002}.

@@

Most implementations of terrain following layers use a coordinate system that makes the computational domain rectangular, but introduces metric terms into the equations of motion.  Instead, the model employed in this study uses Cartesian coordinates and non-orthogonal grids.  By doing so, results from the same model can be compared between terrain following and cut cell grids without modifying the equation set or discretisation.

@@

To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi\), along the edges from \(\Psi\) stored at cell vertices.  Since \(u = \partial \Psi / \partial z\), the streamfunction is found by vertical integration of the velocity profile:
\begin{align}
	\Psi(z) &= \frac{u_0}{2} \left\{ \begin{array}{l l}
		\left( 2z - z_1 - z_2 \right) & \enskip \text{if $z > z_2$} \\
		z - z_1 - \frac{z_2 - z_1}{\pi} \sin \left(\pi \frac{z - z_1}{z_2-z_1}\right) & \enskip \text{if $z_1 < z \leq z_2$} \\
		0 & \enskip \text{if $z \leq z_1$}
	\end{array} \right.
\end{align}

@@

The spatial domain, mountain profile, initial tracer profile and discretisation are the same as those in the horizontal tracer advection test.  The velocity field is defined using a streamfunction, $\Psi$, so that the continuous velocity field is non-divergent and follows the BTF coordinate surfaces given by equation~\ref{eqn:btf} such that
\begin{equation}
	\Psi(x,z) = u_0 H \frac{z - h}{H - h}
\end{equation}
where $u_0 = \SI{10}{\meter\per\second}$, which is the horizontal wind speed where $h(x) = 0$.
The horizontal and vertical components of velocity, $u$ and $w$, are then given by
\begin{align}
	u &= \frac{\partial \Psi}{\partial z} = u_0 \frac{H}{H - h}, \quad w = -\frac{\partial \Psi}{\partial x} = u_0 H \frac{\mathrm{d} h}{\mathrm{d} x} \frac{H - z}{\left( H - h \right)^2}, \nonumber \\
	\frac{\mathrm{d} h}{\mathrm{d} x} &= - h_0 \left[ 
		\beta \cos^2 \left( \alpha x \right) \sin \left( 2 \beta x \right) +
		\alpha \cos^2 \left( \beta x \right) \sin \left( 2 \alpha x \right)
	\right]
\end{align}
Unlike the horizontal advection test, flow extends from the top of the domain all the way to the ground.  The discrete velocity field is calculated using the streamfunction in the same way as the horizontal advection test.

@@

The test setup follows the specification by \cite{klemp2011}.  The domain is \SI{200}{\kilo\meter} wide and \SI{20}{\kilo\meter} high, and the grid resolution is \(\Delta x = \Delta z^\star = \SI{500}{\meter}\).  All boundary conditions are no normal flow.

@@

Following \citet{melvin2010}, the domain is \SI{300}{\kilo\meter} wide and \SI{30}{\kilo\meter} high with a grid resolution of \(\Delta x = \SI{500}{\meter}\) and \(\Delta z^\star = \SI{300}{\meter}\).  \TODO{we want to consider dz=250m and dz=200m, too} The mountain profile has the same form as equation~\ref{eqn:resting:mountain}.  Tests are performed with mountain heights of $h_0 = \SI{250}{\meter}$ and \(h_0 = \SI{500}{\meter}\).  As in the resting atmosphere test, $a = \SI{5}{\kilo\meter}$ is the mountain half-width and $\lambda = \SI{4}{\kilo\meter}$ is the wavelength.  On the optimised SLEVE grid, $s_1 = \SI{5}{\kilo\meter}$ is the large scale height, $s_2 = \SI{2}{\kilo\meter}$ is the small scale height and the optimal exponent value $n = 1.35$ is used.
